Supplementary methods

### 2020-08-25 text/abstract.Rmd

Cancer is a complex disease involving dysfunctional cellular metabolism, other
basic biochemical processes, and the microenvironment. However, in spite of some
generally agreed upon unifying principles [@weinberg2000; @hallmarks_II],
molecular signatures remain largely indistinguishable from tissue-of-origin,
thus obscuring opportunities for novel mechanistic insights and potential
treatment vulnerabilities that hold true across different cancer types. To
address this challenge, we reduced mutation data to disruptions in a select set
of pathways relevant to basic cell biology, from DNA replication to cellular
communication. Using dimensionality reduction techniques, we assigned tumor
samples into ten clusters distinct from their tissues-of-origin and largely free
of bias from mutational burden or clinical stage. We showed that these tumor
classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identify cluster-specific mutations in different
tissues, demonstrating that tissue-specific signatures contribute to common
cellular phenotypes. Moreover, driver gene mutations involved in replication
fidelity and genome stability, such as _BRCA1_ and _Rb1_, are equally distributed
among our clusters, contrary to the expectation that such genes are avatars of
molecular subtype. We also investigated pathways that distinguish between
metastatic and non-metastatic tumors, and found that most meaningful differences
are cluster-specific. Some metastasis enriched pathways from one cluster are
cluster-specific pathways from non-metastatic tumors of another cluster,
suggesting phenotypic convergence, and potentially highlighting key pathways to
focus on for prevention of cluster-specific relapse. Taken as a whole, our
observations suggest that common driver genes combine with tissue-specific
disruptions in complementary sets of tumor-promoting pathways to produce a
limited number of distinct molecular phenotypes. Eventually these signatures
converge on one or more end-stage disease phenotypes. Thus, we present a
coherent view of global tumor biology, and explain how common cellular
dysfunction might arise from tissue-specific mutations.

>>>>

Cancer is a complex disease involving dysfunctional cellular metabolism, other
basic biochemical processes, and the microenvironment. However, in spite of some
generally agreed upon unifying principles [@weinberg2000; @hallmarks_II],
molecular signatures remain largely indistinguishable from tissue-of-origin,
presenting a major barrier for precision health and individualized medicine. To
address this challenge, we reduced mutation data to disruptions in a select set
of pathways relevant to basic cell biology, from DNA replication to cellular
communication. Using dimensionality reduction techniques, we assigned tumor
samples into ten clusters distinct from their tissues-of-origin and largely free
of bias from mutational burden or clinical stage. We showed that these tumor
classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identify cluster-specific mutations in different
tissues, demonstrating that tissue-specific signatures contribute to common
cellular phenotypes. Moreover, driver gene mutations involved in replication
fidelity and genome stability, such as _BRCA1_ and _Rb1_, are equally
distributed among our clusters, contrary to the expectation that such genes are
avatars of molecular subtype. We also investigated pathways that distinguish
between metastatic and non-metastatic tumors, and found that most meaningful
differences are cluster-specific. Some metastasis enriched pathways from one
cluster are cluster-specific pathways from non-metastatic tumors of another
cluster, suggesting phenotypic convergence, and potentially highlighting key
pathways to focus on for prevention of cluster-specific relapse. Taken as a
whole, our observations suggest that common driver genes combine with
tissue-specific disruptions in tumor-promoting pathways to produce a limited
number of distinct molecular phenotypes. Eventually these signatures converge on
one or more end-stage disease phenotypes. Thus, we present a coherent view of
global tumor biology, and explain how common cellular dysfunction might arise
from tissue-specific mutations.

### 2020-08-25 text/survival.Rmd

The degree to which a given cancer accelerates mortality can be determined by
the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier
(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.

>>>>

The degree to which a given cancer accelerates aging can be determined by
the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier
(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.
